# Course Project 1

## load neccessary libraries for performing the anlysis

```{r libraries}
library(ggplot2)
library(plyr)
library(lattice)
```

## read in data from activity.cvs file available from course website and process and transform data into a suitable format

```{r readdata}
# set working directory
setwd("~/MOOC - Data Science/Part 5/Course Project 1")

# read in data
data <- read.csv("activity.csv", header = TRUE, na.strings = "NA")

# format date and interval fields
data$date <- as.POSIXct(data$date, format = "%Y-%m-%d")

# create a file with no na values for steps
nonadata <- data[!is.na(data$steps),]

```

## computing mean total number of steps taken per day (ignoring na valules)

```{r mean}
# calculate total number of steps per day
totalsteps <- aggregate(steps ~ date, data = nonadata, FUN = sum)

# create a histogram of the total number of steps taken each day
hist(totalsteps$steps, xlab = "Steps", main = "histogram of the total number
     of steps taken each day")

# calculate and report the mean and median of the total number of steps taken
# per day

daymean <- mean(totalsteps$steps)
daymean

daymedian <- median(totalsteps$steps)
daymedian

```

### mean of the total number of steps taken per day is equal to `r as.integer(daymean)` and median of the total number of steps taken per dat is equal to `r as.integer(daymedian)`

# calculating the average daily activity pattern

```{r}
# calculate mean number of steps per interval
meansteps <- aggregate(steps ~ interval, data = nonadata, FUN = mean)

# plot the time series
with(meansteps, 
        plot(interval, steps, type = "l",
             xlab = "5-minute interval", ylab = "mean number of steps",
             main = "the average daily activity pattern")
)

# find 5-minute interval, on average across all the days in the dataset, 
# contains the maximum number of steps

intwmaxs <- meansteps[meansteps$steps == max(meansteps$steps),]
intwmaxs

```

### interval `r intwmaxs[[1,1]]` has the maximum (i.e., `r as.integer(intwmaxs[[1,2]])`) number of average steps per day

## imputing missing values

```{r countnas}
nalogic <- ifelse(is.na(data$steps), TRUE, FALSE)
countna <- table(nalogic)[[2]]
```

### there are `r countna` missing values in the data set

### the strategy for filling in all of the missing values in the dataset is to use the mean for that 5-minute interval

```{r impute}
# create a new column to populate with mean of steps in that interval if na
data$newsteps <- ifelse(is.na(data$steps), meansteps[meansteps$interval == data$interval,][[2]], 
                        data$steps)

# calculate total number of steps per day using the imputed dataset
totalnewsteps <- aggregate(newsteps ~ date, data = data, FUN = sum)

# create a histogram of the total number of steps taken each day
hist(totalnewsteps$newsteps, xlab = "New Steps", main = "histogram of the total number
     of steps taken each day using imputed data")

# calculate and report the mean and median of the total number of steps taken
# per day using the imputed dataset

daymeann <- mean(totalnewsteps$newsteps)
daymeann

daymediann <- median(totalnewsteps$newsteps)
daymediann
```

### using imputed dataset, mean of the total number of steps taken per day is equal to `r as.integer(daymeann)` and median of the total number of steps taken per dat is equal to `r as.integer(daymediann)`. These values are diffeerent but close to the previous values where data was not imputed.

## differences in activity patterns between weekdays and weekends
```{r pattern}
# create a new factor variable in the dataset with two levels - "weekday" and
# weekend" indicating whether a given date is a weekday or weekend day
data$day <- as.factor(ifelse(weekdays(data$date) %in% c("Sunday", "Saturday"),
                             "Weekend", "Weekday"))

# create a new dataset for the plot
daydata <- aggregate(newsteps ~ interval + day, data = data, FUN = mean)

# panel plot containing a time series plot of the 5-minute interval
# (x-axis) and the average number of steps taken, averaged across
# all weekday days or weekend days (y-axis)
xyplot(newsteps ~ interval | day, data = daydata, type = "l", layout = c(1,2),
       xlab = "interval", ylab = "average numbe of steps", 
       main = "time series of the 5-minute interval by type of day")
```

### the plots above show that the patterns of average steps of the intervals for weekdays and weekends are different.









